---
title: "Testing power of a Mann-Whitney U-test"
author: "Katarzyna Sidorczuk"
date: "4/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(biogram)
library(dplyr)
library(ggplot2)

generate_sequences <- function(n_seq, alphabet, probs, min_len = 50, max_len = 100) {
  lapply(1:n_seq, function(i) {
    len <- sample(min_len:max_len, 1)
    sample(alphabet, len, replace = TRUE, prob = probs)
  })
}



test_power <- function(fun_name, aa, aa_comp1, aa_comp2, dataset_size, p_adjustment = "BH", ...) {
  seqs1 <- generate_sequences(dataset_size, aa, aa_comp_1) %>% 
    setNames(paste0("Set1_", 1:dataset_size))
  seqs2 <- generate_sequences(dataset_size, aa, aa_comp2) %>% 
    setNames(paste0("Set2_", 1:dataset_size))
  datasets <- list("Dataset1" = seqs1, "Dataset2" = seqs2)
  aa_comp <- lapply(names(datasets), function(ith_set) {
    lapply(names(datasets[[ith_set]]), function(ith_prot) {
      aa <- unlist(datasets[[ith_set]][[ith_prot]], use.names = FALSE)
      data.frame(table(factor(aa, levels = c("A", "C", "D", "E", "F", "G", "H", 
                                             "I", "K", "L", "M", "N", "P", "Q", 
                                             "R", "S", "T", "V", "W", "Y")))/length(aa)) %>% 
        setNames(c("Amino acid", "Frequency"))  %>% 
        mutate(seq_name = ith_prot,
               Dataset = ith_set)
    }) %>% bind_rows()
  }) %>% bind_rows()
  
  lapply(toupper(colnames(biogram::aaprop)), function(ith_aa) {
    data.frame(
      aa = ith_aa,
      pval = match.fun(fun_name)(x = filter(aa_comp, Dataset == "Dataset1" & `Amino acid` == ith_aa)[["Frequency"]],
                                 y = filter(aa_comp, Dataset == "Dataset2" & `Amino acid` == ith_aa)[["Frequency"]],
                                 ...)[["p.value"]]
    )
  }) %>% bind_rows() %>% 
    mutate(pval_adjusted = p.adjust(pval, p_adjustment))
}

```

### Amino acid distributions used in the simulation

Three distributions of amino acids were used for testing:

- ```Dataset_1``` - all amino acids with equal probabilities
- ```Dataset_2``` - various probabilities of amino acid occurences (very high for some and very low for others)
- ```Dataset_3``` - small differences between amino acid probabilities

The default *p*-value adjustment method is Benjamini-Hochberg correction. 

```{r}
aa <- toupper(colnames(biogram::aaprop))

aa_comp_1 <- rep(0.05, 20)
aa_comp_2 <- c(rep(0.1, 2), rep(0.01, 6), rep(0.025, 3), 0.3, rep(0.08, 3), rep(0.04, 3), rep(0.0025, 2))
aa_comp_3 <- c(rep(0.06, 5), rep(0.05, 10), rep(0.04, 5))

```

### Equal probabilities vs. large differences in probabilities, 1000 peptides

```{r}
test_power("wilcox.test", aa, aa_comp_1, aa_comp_2, 1000, exact = FALSE)
```

### Equal probabilities vs. large differences in probabilities, 4000 peptides
```{r}
test_power("wilcox.test", aa, aa_comp_1, aa_comp_2, 4000, exact = FALSE)
```

### Equal probabilities vs. small differences in probabilities, 1000 peptides
```{r}
test_power("wilcox.test", aa, aa_comp_1, aa_comp_3, 1000, exact = FALSE)
```

### Equal probabilities vs. equal probabilities, 1000 peptides
```{r}
test_power("wilcox.test", aa, aa_comp_1, aa_comp_1, 1000, exact = FALSE)
```

### Equal probabilities vs. small differences in probabilities, 1000 peptides, Bonferroni adjustment
```{r}
test_power("wilcox.test", aa, aa_comp_1, aa_comp_2, 1000, exact = FALSE, adjustment = "bonferroni")
```


### Results of tests on all three datasets with 4000 peptides

```{r echo = FALSE, message = FALSE, fig.height=3, fig.width = 10}
seqs1 <- generate_sequences(4000, aa, aa_comp_1)
seqs2 <- generate_sequences(4000, aa, aa_comp_2)
seqs3 <- generate_sequences(4000, aa, aa_comp_3)

datasets <- list("Dataset1" = seqs1, "Dataset2" = seqs2, "Dataset3" = seqs3)
aa_comp <- lapply(names(datasets), function(ith_set) {
  lapply(datasets[[ith_set]], function(ith_prot) {
    aa <- unlist(ith_prot, use.names = FALSE)
    data.frame(table(factor(aa, levels = c("A", "C", "D", "E", "F", "G", "H", 
                                           "I", "K", "L", "M", "N", "P", "Q", 
                                           "R", "S", "T", "V", "W", "Y")))/length(aa)) %>% 
      setNames(c("Amino acid", "Frequency"))  %>% 
      mutate(Dataset = ith_set)
  }) %>% bind_rows()
}) %>% bind_rows()


combns <- combn(unique(aa_comp[["Dataset"]]), 2, simplify = FALSE)

test_res <- lapply(seq_along(combns), function(ith_combn) {
  test_dat <- filter(aa_comp, Dataset %in% combns[[ith_combn]])
  lapply(unique(aa_comp[["Amino acid"]]), function(ith_aa) {
    data.frame(comparison = paste0(combns[[ith_combn]][1], "_", combns[[ith_combn]][2]),
               aa = ith_aa,
               pval = wilcox.test(x = filter(test_dat, `Amino acid` == ith_aa, Dataset == combns[[ith_combn]][1])[["Frequency"]],
                                  y = filter(test_dat, `Amino acid` == ith_aa, Dataset == combns[[ith_combn]][2])[["Frequency"]],
                                  exact = FALSE)[["p.value"]]) 
  }) %>% bind_rows() %>% 
    mutate(pval_adjusted = p.adjust(pval))
}) %>% bind_rows()


test_res %>% 
  select(-pval) %>% 
  group_by(aa, comparison) %>% 
  summarise(is_significant = as.logical(pval_adjusted < 0.05)) %>% 
  ggplot(aes(x = aa, y = comparison, fill = is_significant)) +
  geom_tile(color = "white") +
  theme_bw()
```

The above simulations show that Mann-Whitney U-test displays high sensitivity in detection of differences in amino acid composition, especially in large datasets. 
